Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-02-15 paper/letter.tex
main body updated

Recent models computational models \cite{GaravitoCamargo2014} have
shown that rotation also imprints an effect on the Lyman-alpha
morphology.  
The most important consequence of rotation is that spherical 
symmetry is broken.
The line morphology now depends on the viewing angle respect to the
rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity.  
When the rotational velocity is close to the half-line width of the
static line, the line becomes single peaked as it is observed in
\tol, a unique feature that other theoretical models find
impossible to reproduce.

\tol is a compact star forming dwarf galaxies (CDGs) that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two features that had
so far evaded a natural physical interpratation.
First, it has symmetric profile  without any systemic displacement
from the expected line-center. 
Second, it shows a triple peak, while the most common observational
feature corresponds either to a single or asymetric double peak. 
We have attempted an interpretation of the Ly-$\alpha$ line in
\tol\ by including the effect of gas bulk rotation.
We find that this model provides a satisfactory explanation of \tol's
puzzling features.

Dots in Figure 1. represebt the observational data for \tol with the
overplot from our best fit model from the full radiative transfer
simulation.   The parameters for the best fit are
$v_{max}=300$\kms, $\tau=1\times10^7$, $T=1.5\times 10^{4}$K.
The optical depth and temperature can be translated into a column
density of $N_{HI}$. Our model is also able to constrain the angle
between the rotation axis and the observational line-of-sight to $\theta=65^{\circ}\pm$.
To constrain the model uncertainties we perform a Markov Chain Monte
Carlo calculation using an approximate analytical solution for the
lya spectrum in the bulk rotation case.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{tolfit.png}
\caption{Nice fit}
\end{center}
\end{figure}

Radio surveys have put an upper limit to the neutral hydrogen mass in
\tol of $M<2.65\times 10^{8}$ M$_{\odot}$. Together with our velocity
and column density results, under the assumption of spherical
symmetry, we can constrain the total dynamical mass of the HI region
to be in the range $2.1\times 10^{9}$M$_{\odot}$ $<M_D<$  $6.6\times
10^{9}$M$_{\odot}$ and the diameter of HI region should
have a diameter of $0.11 < D/\mathrm{kpc}<0.34$. This makes the
dynamical mass  at least $7$ to $25$ times larger than the H$I$ mass.

We consider three possibilities to contribute to this high
dynamical mass: stars, dark matter and a supermassive black
hole. The stellar mass can be constrained by measurements of the
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution . Considering a standard variations for the dark matter
halo, it can only constributio to $xx$ M$_{\odot}$ in the region of
interest. A super massive black hole remains an open
possibility. Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total objects mass, suggesting that
there is a large population of undetected black holes.

A future observational test of this possibility in \tol\ would require
integral field unit measurements spatially resolving its spatial
extent. 
\tol\ spans a region of $\approx 4$ arcminutes. 
An instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a detailed mapping of different
ionization lines and infer a kinematic map.

Why are not more examples of triple peaked emission in observed
Lyman-alpha emitting gaalaxies?  
The ubiuquity of galaxy outflows overimposed to the rotation
transforms the line into the common double or single peaked
observations.  
Therefore, the absence of HI outflows in \tol\ is another special
feature that permitted a symmetric triple peaked lya line to emerge.

The implications of our finding for the study of lyman-alpha
emitters, including the first generation of galaxies are
manifold. First of all, it provides support to including rotation as a
feature to interpret observed spectra. Secondly, it provides a
motivation to explore under what conditions rotation dominated
dynamics are expected in comparison with outflow dominated
dynamics. Finally, if the hypothesis of a supermassive black hole in
\tol\ proves to be consistend with future observational kinematic
maps, it would provide a new information to test and probe theories on
the coevolution of galaxies and black holes.

>>>>

Recent computational models \cite{GaravitoCamargo2014} have
shown that rotation also imprints an effect on the Lyman-alpha
morphology.  
The most important consequence of rotation is that spherical 
symmetry is broken.
The line morphology now depends on the viewing angle respect to the
rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity.  
When the rotational velocity is close to the half-line width of the
static line, the line becomes single peaked as it is observed in
\tol, a unique feature that other theoretical models find
impossible to reproduce.

\tol\ is a compact star forming dwarf galaxies (CDGs) that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two features that had
so far evaded a natural physical interpratation.
First, it has symmetric profile  without any systemic displacement
from the expected line-center. 
Second, it shows a triple peak, while the most common observational
feature corresponds either to a single or asymetric double peak. 
We have attempted an interpretation of the Ly-$\alpha$ line in
\tol\ by including the effect of gas bulk rotation.
We find that this model provides a satisfactory explanation of \tol's
puzzling features.

Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit model from the full radiative transfer
simulation.   
The parameters for the best fit are
$v_{max}=300$\kms, $\tau=1\times10^7$, $T=1.5\times 10^{4}$K.
The optical depth and temperature can be translated into a column
density of $\log N_{HI} / \mathrm{atoms\ cm}^{-2} =
  20.5$.   
Our model is also able to constrain the angle
between the rotation axis and the observational line-of-sight to $\theta=65^{\circ}\pm$.
To determine the uncertainties we perform a Markov Chain Monte
Carlo calculation using an approximate analytical solution for the
Ly-$\alpha$ spectrum including rotational effects.

Radio surveys of the 21cm line have put an upper limit to the neutral
hydrogen mass in \tol of $M<2.65\times 10^{8}$ M$_{\odot}$. 
Together with our velocity and column density results, under the
assumption of spherical symmetry, we can constrain the total dynamical
mass of the HI region to be in the range $2.1\times 10^{9}$M$_{\odot}$
$<M_D<$  $6.6\times 10^{9}$M$_{\odot}$ and the diameter of HI region
should have a diameter of $0.11 < D/\mathrm{kpc}<0.34$. 
This makes the
dynamical mass  at least $7$ to $25$ times larger than the H$I$ mass.

We consider three possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass can be constrained by measurements of the 
stellar continuum, this can only represent $xx$M$_{\odot}$ of the total
contribution. 
Considering a standard dark matter halo, it can only constributio to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open  possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

A future observational test of this possibility in \tol\ would require
integral field unit measurements spatially resolving its spatial
extent. 
\tol\ spans a region of $\approx 4$ arcminutes. 
An instrument such as the Multi Unit Spectroscopic Explorer with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a detailed mapping of different
ionization lines to infer a kinematic map.

Why is triple peaked emission not common in  Lyman-alpha emitting
gaalaxies?   
The ubiuquity of galaxy outflows overimposed to the rotation
transforms the line into the common double or single peaked
line.
Therefore, the absence of HI outflows in \tol\ is another special
feature that permitted a symmetric triple peaked lya line to emerge.

The implications of our finding for the study of LAEs, including the
first generation of galaxies are manifold. 
First of all, it demonstrated the importance of including rotation as
theoretical feature to interpret observed spectra. 
Secondly, it provides a motivation to explore under what conditions
rotation dominated dynamics are expected in comparison with outflow
dominated dynamics. 
Finally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistend with future observational
kinematic maps, it would provide a new information to test and probe
theories on the coevolution of galaxies and black holes in the first
generation of galaxies.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{tolfit.png}
\caption{{\bf Symmetric triple peaked Ly-$\alpha$ emission of \tol.}
  Dots correspond to the observational data. The line shows the result
of our best model from a full radiative transfer simulation. The best
parameters correspond to a rotational velocity of $300$km s$^{-1}$,
neutral Hydrogen optical depth of $10^{7}$ and a viewing angle of
$65^{\circ}$ measured from the rotation axis.} 
\end{center}
\end{figure}
